---
title: "Best guess of initial covariance parameters"
author: "Alessandro Samuel Rosa"
date: "30 November 2015"
output: html_document
---

```{r}
rm(list = ls())
require(pedometrics)
require(georob)
```

for four soil variables contained in the meuse dataset
(cadmium, copper, zinc, and lead)

# Estimator of the sample variogram

Our first test regards the selection of the best estimator for computing the
sample variogram. Four options are evaluated:

1. Genton's robust Qn-estimator,
2. Dowd's robust MAD-estimator,
3. non-robust method-of-moments estimator, and
4. robust Cressie-Hawkins estimator.

Evaluation involves comparing the guesses made using each estimator of the
sample variogram. We also evaluate the effect of the number of point-pairs in 
the first lag-distance class (19 or 164). This is possible because the
number of point-pairs in the second lag-distance class is 19 (the number of 
point-pairs in the first lag-distance class is equal to zero). We then fit
geostatistical models using the ***georob***-package to evaluate the
performance of each estimator of the sample variogram. Our goal is to compare 
the guesses with the estimated variogram model parameters to see if the 
estimator of the sample variogram affects the estimated variogram model 
parameters.

## meuse dataset

```{r}
require(sp)
data(meuse)
```

### npairs = 19

We observed that the guessed ranges and sills are approximately the same for all
four estimators, except for two variables (zinc and lead), where "matheron" 
guessed shorter ranges. The main difference is in the guess of the nugget 
variance, and thus the partial sill. In general, "matheron" guesses the nugget
variance to be very high compared to the other three estimator of the sample
variogram. Estimators "qn", "mad", and "ch" seem to produce closer guesses for
the nugget variance. Overall, "qn" produces the "better looking"" sample
variogram, specially close to the origin, what indicates its robustness to
outlying observations. However, all robust estimators of the sample variogram
failed one or more times in producing a "good looking" sample variogram.

Estimated covariance parameters for zinc and lead are considerably different
from the guesses made with "matheron". However, the estimated variogram model
parameters are very close for all four estimators of the sample variogram. The
observed fluctuations do not evidence any trend, and (pehaps) can be regarded 
as a result of fluctuations in the optimizer. As such, all three robust 
estimators of the variogram can be considered satisfactory.

The guesses of the nugget and total variance (sill) were higher using "matheron"
for three variables (cadmium, zinc, and lead) and using "mad" for one variable
(lead).

Best estimator

Variable | Range      | Partial sill | Nugget 
-------- | ---------- | ------------ | ----------
chromium | "pooled"   | "mad"        | "matheron"
copper   | "pooled"   | "matheron"   | "matheron"
zinc     | "matheron" | "mad"        | "mad"
lead     | "pooled"   | "ch"         | "mad"

Worse estimator

Variable | Range      | Partial sill | Nugget 
-------- | ---------- | ------------ | ----------
chromium | "mad"      | "ch"         | "qn"
copper   | "matheron" | "mad"        | "mad"
zinc     | "mad"      | "qn"         | "matheron"
lead     | "qn"       | "matheron"   | "matheron"

### npairs = 164

We observe again that the guessed ranges and sills are
approximately the same for all four estimators, the main difference among
estimators being in the guess of the nugget variance. The major difference from
the previous analysis is that the guess of the nugget variance made by 
"matheron" is much higher. This is more pronounced when the data has a longer
tail. Compared to "matheron", the other three estimators produce similar 
guesses of the nugget variance. This suggests that, as the number of 
point-pairs in the first lag-distance class increases, robust estimators produce
very similar sample variograms. In fact, the relative shape of the sample
variogram estimated using "matheron" is similar to those 
produced using robust estimators. However, because "matheron" is a non-robust
estimator, the presence of outlying observations results in an overestimation
of the (nugget) variance, while robust estimators can underesitmate the (nugget)
variance. This was already observed by Lark (2000) [A comparison of some robust
estimators of the variogram for use in soil survey. European Journal of Soil
Science, v. 51, p. 137-157.].

Best estimator

Variable | Range      | Partial sill | Nugget 
-------- | ---------- | ------------ | ----------
chromium | "matheron" | "matheron"   | "matheron"
copper   | "qn"       | "mad"        | "mad"
zinc     | "matheron" | "mad"        | "mad"
lead     | "pooled"   | "ch"         | "mad"

Worse estimator

Variable | Range      | Partial sill | Nugget 
-------- | ---------- | ------------ | ----------
chromium | "mad"      | "mad"        | "mad"
copper   | "matheron" | "matheron"   | "matheron"
zinc     | "mad"      | "matheron"   | "matheron"
lead     | "mad"      | "matheron"   | "matheron"

Estimator "matheron" produced the best guess of the nugget variance for the data
with a longer tail. The variogram model fitted to this data has a nugget 
variance larger than the partial sill. Robust estimators of the sample variogram
produced guesses that underestimate the nugget variance, what indicates that 
they may be too optimistic regarding the nugget variance when the data is
non-normal. However, when the data had a normal distribution, robust estimators
produced the best guesses of the nugget variance, "matheron" being the poorest
method because it consistently guessed a nugget variance larger that the partial
sill. Despite these differences, the estimated variogram model parameters are 
not considerably different for all four estimators of the sample variogram.

```{r,eval=FALSE}
par(mfrow = c(2, 2))
```

```{r,eval=FALSE}
# Setup data
# meuse$z <- lm(log(cadmium) ~ sqrt(dist), meuse)$resid
# meuse$z <- lm(log(copper) ~ sqrt(dist), meuse)$resid
# meuse$z <- lm(log(zinc) ~ sqrt(dist), meuse)$resid
meuse$z <- lm(log(lead) ~ sqrt(dist), meuse)$resid

# Guess initial covariance parameters
estimator <- c("qn", "mad", "matheron", "ch")
icp <- sapply(
  1:length(estimator), function (i)
    vgmICP(z = meuse$z, coords = meuse[, 1:2], method = "a", plotit = T,
           estimator = estimator[i], min.npairs = 19))
estimator <- c(estimator, "pooled")
icp <- cbind(icp, rowMeans(icp))
colnames(icp) <- estimator
estimator[which.max(icp["nugget", ])]
estimator[which.max(icp["nugget", ] + icp["p_sill", ])]

# Fit variogram model
fit_georob <- list()
for (i in 1:length(estimator)) {
  fit_georob[[i]] <- georob::georob(
    z ~ 1, data = meuse, locations = ~ x + y, variogram.model = "RMexp",
    verbose = 0, tuning.psi = 1000,
    param = c(variance = icp[2, i], nugget = icp[3, i], scale = icp[1, i]))
}

# Check estimated parameters
est_par_georob <- sapply(1:length(estimator), function (i) 
  round(fit_georob[[i]]$param, 4))[c(4, 1, 3),]
colnames(est_par_georob) <- estimator
round(icp, 4) 
est_par_georob
rbind(min = estimator[apply(abs(round(icp, 4) - est_par_georob), 1, which.min)],
      max = estimator[apply(abs(round(icp, 4) - est_par_georob), 1, which.max)])
round(icp, 4) - est_par_georob
```

### Partial conclusion

Robust estimators of the sample variogram produced similar guesses of the 
initial covariance parameters. In most cases, they outperformed the non-robust
estimator of the sample variogram as measured by the closeness between guessed
and estimated variogram model parameters. Because the estimated variogram model
parameters are virtually the same irrespective of the estimator of the sample 
variogram, we can assume that they accurately fit the data. Thus, we can 
conclude that robust estimators of the sample variogram had a better performance
for the `meuse` dataset. Because "qn" generaly produced "better looking" sample 
variograms, specially close to the origin, and because it is the default option
in the ***georob***-package, we can suggest it to be the default estimator of
the sample variogram for guessing the initial covariance parameters.

## jura dataset

```{r,eval=FALSE}
require(gstat)
data(jura)
```

nugget: "matheron" "matheron" "matheron" "matheron"
sill: "mad" "matheron" "matheron" "matheron"

Best estimator

Variable | Range      | Partial sill | Nugget 
-------- | ---------- | ------------ | ----------
Cd       | "qn"       | "mad"        | "mad"
Cu       | "matheron" | "mad"        | "mad" 
Zn       | "pooled"   | "mad"        | "mad"
Pb       | NA         | "mad"        | "qn"      

Worse estimator

Variable | Range      | Partial sill | Nugget 
-------- | ---------- | ------------ | ----------
Cd       | "mad"      | "matheron"   | "matheron"
Cu       | "ch"       | "matheron"   | "matheron"
Zn       | "qn"       | "matheron"   | "matheron"
Pb       | NA         | "matheron"   | "matheron"

```{r,eval=FALSE}
par(mfrow = c(2, 2))
```

```{r,eval=FALSE}
# Setup data
jura.pred$z <- lm(log(Cd) ~ Rock + Landuse, jura.pred)$resid # 1, 2, 4
# jura.pred$z <- lm(log(Cu) ~ Rock + Landuse, jura.pred)$resid # 3
# jura.pred$z <- lm(log(Zn) ~ Rock + Landuse, jura.pred)$resid # 1, 2, 3, 4
# jura.pred$z <- lm(log(Pb) ~ Rock + Landuse, jura.pred)$resid # 3

# Guess initial covariance parameters
estimator <- c("qn", "mad", "matheron", "ch")
icp <- sapply(
  1:length(estimator), function (i)
    vgmICP(z = jura.pred$z, coords = jura.pred[, 1:2], 
           method = "a", plotit = T, 
           estimator = estimator[i]))
estimator <- c(estimator, "pooled")
icp <- cbind(icp, rowMeans(icp))
colnames(icp) <- estimator
estimator[which.max(icp["nugget", ])]
estimator[which.max(icp["nugget", ] + icp["p_sill", ])]

# Fit variogram model
fit_georob <- list()
for (i in 1:length(estimator)) {
  fit_georob[[i]] <- georob::georob(
    z ~ 1, data = jura.pred, locations = ~ Xloc + Yloc, 
    variogram.model = "RMexp", verbose = 0, tuning.psi = 1000,
    param = c(variance = icp[2, i], nugget = icp[3, i], scale = icp[1, i]))
}

# Check estimated parameters
est_par_georob <- sapply(1:length(estimator), function (i) 
  round(fit_georob[[i]]$param, 4))[c(4, 1, 3),]
colnames(est_par_georob) <- estimator
round(icp, 4)
est_par_georob
rbind(min = estimator[apply(abs(round(icp, 4) - est_par_georob), 1, which.min)],
      max = estimator[apply(abs(round(icp, 4) - est_par_georob), 1, which.max)])
round(icp, 4) - est_par_georob
```

## Santa Maria dataset

### min.npairs = 28



```{r,eval=FALSE}
load("~/projects/dnos-sm-rs/data/R/1stArticlePartI.rda")
cal_data <- data.frame(
  cal_data@coords, clay_sel$best_lm$residuals, orca_sel$best_lm$residuals, 
  ecec_sel$best_lm$residuals)
colnames(cal_data) <- c("x", "y", "clay", "orca", "ecec")
```

```{r, eval=FALSE}
par(mfrow = c(2, 2))
```

```{r,eval=FALSE}
# Setup data
cal_data$z <- lm(clay ~ 1, cal_data)$resid # 1, 2, 4
# cal_data$z <- lm(orca ~ 1, cal_data)$resid # 3
# cal_data$z <- lm(ecec ~ 1, cal_data)$resid # 1, 2, 3, 4

# Guess initial covariance parameters
estimator <- c("qn", "mad", "matheron", "ch")
icp <- sapply(
  1:length(estimator), function (i)
    vgmICP(z = cal_data$z, coords = cal_data[, 1:2], method = "a", plotit = T, 
           estimator = estimator[i], min.npairs = 30))
estimator <- c(estimator, "pooled")
icp <- cbind(icp, rowMeans(icp))
colnames(icp) <- estimator
estimator[which.max(icp["nugget", ])]
estimator[which.max(icp["nugget", ] + icp["p_sill", ])]

# Fit variogram model
fit_georob <- list()
for (i in 1:length(estimator)) {
  fit_georob[[i]] <- georob::georob(
    z ~ 1, data = cal_data, locations = ~ x + y, 
    variogram.model = "RMexp", verbose = 0, tuning.psi = 1,
    param = c(variance = icp[2, i], nugget = icp[3, i], scale = icp[1, i]))
}

# Check estimated parameters
est_par_georob <- sapply(1:length(estimator), function (i) 
  round(fit_georob[[i]]$param, 4))[c(4, 1, 3),]
colnames(est_par_georob) <- estimator
round(icp, 4)
est_par_georob
rbind(min = estimator[apply(abs(round(icp, 4) - est_par_georob), 1, which.min)],
      max = estimator[apply(abs(round(icp, 4) - est_par_georob), 1, which.max)])
round(icp, 4) - est_par_georob
```

# 2) Method used to make the initial guess #####################################
# We now test five method to make the initial guess:
# a) Samuel-Rosa,
# b) JianEtAl1996,
# c) HiemstraEtAl2009,
# d) DesassisEtAl2012, and
# e) LarrondoEtAl2003.
# Methods a and c use exponentially spaced lag classes, while methods b, d and e
# use equidistant lag classes. Evaluation involves comparing the guesses made 
# using each method for four soil variables contained in the meuse dataset
# (cadmium, copper, zinc, and lead).
# 
# The results show that methods a, b, and c result in (virtually) the same
# estimated of the covariance parameters, although producing different initial
# guesses. Methods d and e result in errors during execution. These errors 
# result from the nugget being estimated as zero. I do not know how georob
# handles these errors, but the estimated covariance parameters usually are
# significantly different from those estimated using methods a, b, and c. The
# largest differences appear when the data has a longer tail.
# 
# Initial guesses provided by methods a and c seem to be closer to the estimated
# covariance paramenters. Because method a is more elaborated, we follow with 
# it in the next analysis.

```{r,eval=FALSE}
# meuse$z <- lm(log(cadmium) ~ sqrt(dist), meuse)$resid;hist(meuse$z)
# meuse$z <- lm(log(copper) ~ sqrt(dist), meuse)$resid;hist(meuse$z)
# meuse$z <- lm(log(zinc) ~ sqrt(dist), meuse)$resid;hist(meuse$z)
# meuse$z <- lm(log(lead) ~ sqrt(dist), meuse)$resid;hist(meuse$z)
method <- letters[1:5]
icp <- sapply(1:length(method), function (i) 
  vgmICP(z = meuse$z, coords = meuse[, c("x", "y")], method = method[i], 
         cov.model = "Exp", min.npairs = 20))
colnames(icp) <- method
fit_georob <- lapply(1:length(method), function (i)
  tmp <- georob::georob(
    z ~ 1, data = meuse, locations = ~ x + y, variogram.model = "RMexp",
    verbose = 2, tuning.psi = 2,
    param = c(variance = icp[2, i], nugget = icp[3, i], scale = icp[1, i])))
est_par_georob <- sapply(
  1:length(method), function (i) round(fit_georob[[i]]$param, 4))[c(4, 1, 3),]
colnames(est_par_georob) <- method
round(icp, 4)
est_par_georob
```

# 3) Evaluate the fitting function #############################################
# We evaluate three fitting functions:
# a) georob:georob,
# b) geoR::likfit, and
# c) gstat::fit.variogram.
# The analysis includes comparing the estimated covariance parameters, and how
# close the initial guess is to that.

```{r,eval=FALSE}
# meuse$z <- lm(log(cadmium) ~ sqrt(dist), meuse)$resid;hist(meuse$z)
# meuse$z <- lm(log(copper) ~ sqrt(dist), meuse)$resid;hist(meuse$z)
# meuse$z <- lm(log(zinc) ~ sqrt(dist), meuse)$resid;hist(meuse$z)
# meuse$z <- lm(log(lead) ~ sqrt(dist), meuse)$resid;hist(meuse$z)
geodata <- list(coords = meuse[, 1:2], data = meuse$z)
vario <- gstat::variogram(
  z ~ 1, data = meuse, locations = ~ x + y, 
  boundaries = pedometrics::vgmLags(meuse[, 1:2]))
icp <- unname(vgmICP(z = meuse$z, coords = meuse[, 1:2], cov.model = "Exp"))
fit_georob <- georob::georob(
    z ~ 1, data = meuse, locations = ~ x + y, variogram.model = "RMexp",
    verbose = 2, tuning.psi = 2, 
    param = c(variance = icp[2], nugget = icp[3], scale = icp[1]))
fit_geor <- geoR::likfit(
  geodata = geodata, lik.method = "REML", 
  ini.cov.par = c(icp[2], icp[1]), nugget = icp[3])
fit_gstat <- gstat::fit.variogram(
  vario, gstat::vgm(psill = icp[2], "Exp", range = icp[1], nugget = icp[3]))

# Compare estimated covariance parameters
est_par_georob <- round(fit_georob$param[c(3, 1, 4)], 3)
est_par_geor <- round(c(fit_geor$nugget, fit_geor$cov.pars), 3)
est_par_gstat <- round(
  c(fit_gstat$psill[1], fit_gstat$psill[2], fit_gstat$range[2]), 4)
data.frame(
  icp = round(rev(icp), 3), gstat = est_par_gstat, geor = est_par_geor, 
  georob = est_par_georob, nongeorob = est_par_georob_non)
# Plot variogram lines
v <- georob::sample.variogram(
  meuse$z, locations = meuse[, 1:2], 
  lag.dist.def = pedometrics::vgmLags(meuse[, 1:2], 5))
plot(v$lag.dist, v$gamma, ylim = c(0, max(v$gamma) * 1.5))
lines(fit_georob)
lines(fit_geor, col = "blue", max.dist = max(v$lag.dist))
v <- gstat::variogramLine(fit_gstat, dist_vector = vario$dist)
lines(spline(v$dist, v$gamma, method = "natural", n = 101), col = "green")
icp_vgm <- gstat::variogramLine(
  gstat::vgm(psill = icp[2], "Exp", range = icp[1], nugget = icp[3]),
  dist_vector = vario$dist)
lines(spline(icp_vgm$dist, icp_vgm$gamma, method = "natural", n = 101), 
      col = "red")
# Cross-validation
cv_georob <- georob::cv(
  fit_georob, method = "random", seed = 2001, ncores = 1, nset = 155,
  verbose = 4, re.estimate = FALSE)
set.seed(2001)
cv_gstat <- gstat::krige.cv(
  formula = z ~ 1, data = meuse, fit_gstat, nfold = 155, locations = ~ x + y)
set.seed(2001)
cv_geor <- xvalid(geodata, model = fit_geor, locations.xvalid = "all")
cv_georob <- cv_georob$pred$pred - meuse$z
cv_gstat <- cv_gstat$var1.pred - meuse$z
cv_geor <- cv_geor$predicted - meuse$z
data.frame(
  me = apply(data.frame(cv_georob, cv_gstat, cv_geor), 2, mean),
  mse = apply(data.frame(cv_georob^2, cv_gstat^2, cv_geor^2), 2, mean),
  mae = apply(data.frame(abs(cv_georob), abs(cv_gstat), abs(cv_geor)), 2, mean))


```

